Machine Learning Methods for Discourse Segmentation of Communications in E-Mail Based Behavioral Interventions

### 2018-03-09 paper-amia-2018/manuscript.tex
Alex edited

\section*{Results}
Experimental results are evaluated with ``boundary'' and ``not boundary'' classes as well as their weighted average, which are shown in Table~\ref{tab:result_boundary}, \ref{tab:result_not_boundary}, and \ref{tab:result_weighted_avg}, respectively.\\

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM, KNN, and RNN methods for detecting segmentation boundary in e-coaching text. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_boundary}
  \begin{tabular}{|l|l|l|l|p{0.15\linewidth}|p{0.15\linewidth}|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{lexical features only}} & \multicolumn{3}{|c|}{\textbf{lexical + punctuation marks (+topic-based except RNN)}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-measure} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-measure}\\ \hline

@@

As follows from Table~\ref{tab:result_boundary}, KNN performs best among all machine learning models in terms of precision and F1-measure, achieved 0.808 precision with 0.728 F1-measure when lexical features are used; and 0.820 precision with 0.779 F1-measure when a combination of lexical, punctuation mark and topic model-based features are used. However, NB demonstrates the lowest performance among all models in terms of all performance metrics. In this study, GRU appears as the second highest model, obtains 10.68\% higher recall with 3.78\%  higher F1-measure than LSTM. On the other hand, SVM exhibits highest recall 0.679 when only lexical features are used. When lexical features are used in combination with punctuation mark and topic model-based features, recall increases by 0.6\%, 2.5\%, and 11.92\%; and F1-measure increases by 0\%, 3.39\%, and 7\% for NB, SVM, and KNN models, respectively. Nevertheless, precision increases by 4.31\% and 1.49\% for SVM and KNN methods while decreases by 0.7\% in NB. \\

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM, KNN, and RNN methods for the identification of ``not boundary'' class. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_not_boundary}
  \begin{tabular}{|l|l|l|l|p{0.15\linewidth}|p{0.15\linewidth}|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{lexical features only}} & \multicolumn{3}{|c|}{\textbf{lexical + punctuation marks (+ topic-based except RNN)}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-measure} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-measure}\\ \hline

@@

Table~\ref{tab:result_not_boundary} summarizes the performance of NB, SVM, KNN, and RNN models for detecting ``not boundary'' class in e-coaching text. We observed that performance is remarkably high in ``not boundary'' class compared to boundary detection which is expected because 96.71\% instances are from ``not boundary'' class. KNN achieves 22.40\%, 50.07\%, and 36.26\% higher precision, recall, and F1-measure, respectively, for lexical features; and 20.85\%, 33.96\%, and 27.47\% higher precision, recall, and F1-measure, respectively, for combined features compared to boundary class. In contrast to boundary detection, RNN demonstrates the highest performance among all models. LSTM obtains 0.993 precision with 0.994 recall and F1-measure while GRU exhibits 0.994 for all performance metrics. Impact of punctuation mark and topic model-based features is also consistent with ``not boundary'' classification. Results show that F1-measure increases by 0\%, and 0.1\% for SVM and KNN models although decreases by 0.1\% for NB. \\

\begin{table}[ht]
\centering
\caption{Weighted average performance of NB, SVM, KNN, and RNN methods for the segmentation of e-coaching text in detecting both ``boundary'' and ``not boundary'' classes. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_weighted_avg}
  \begin{tabular}{|l|l|l|l|p{0.15\linewidth}|p{0.15\linewidth}|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{lexical features only}} & \multicolumn{3}{|c|}{\textbf{lexical + punctuation marks (+ topic-based except RNN)}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-measure} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-measure}\\ \hline

@@

Table~\ref{tab:result_weighted_avg} outlines the weighted average results of the experiment on the models for the segmentation of e-coaching text by classifying them into ``boundary'' and ``not boundary'' classes. Overall, KNN obtains the best performance with all metrics and NB denotes the lowest performance among all methods. GRU demonstrates the same result as KNN for precision and F1-measure except for recall. SVM shows moderate performance, obtains 0.981, 0.982, and 0.983 for precision, recall, and F1-measure, respectively, when lexical features are used. Influence of the additional features is also consistent as mentioned in Table~\ref{tab:result_boundary} and~\ref{tab:result_not_boundary}. Precision increases by 0.1\%, 0.2\%, and 0.3\%; recall increases by 0\%, 0.1\%, and 0.2\%; and F1-measure increases by 0\%, 0.2\%, and 0.3\% for NB, SVM, and KNN methods, respectively, when combined features are used.\\

\section*{Discussion}
This study is the first efforts to evaluate the automatic segmentation of e-coaching text. Experimental results indicate that KNN is the best model among all machine learning methods considered for this study. KNN achieved 0.986 F1-measure in overall, 0.779 and 0.993 F1-measures for detecting ``boundary'' and ``not boundary'', respectively. The robust performance of KNN provides the evidence that machine learning models are capable to learn information from clinical exchange. Although the domain of this study was intentionally quite small, we believe that our study is not limited to the e-coaching domain, and it can be successfully applied to other domain as well.

Punctuation mark and topic model-based features made a significant improvement in performance of all machine learning methods. Nearly all cases, the model performs better when lexical features are used in combination with punctuation mark and topic model-based features. This results also mean that segmentation performance might be improved by adding more relevant features including human insight into the problem.

@@

In this paper, results are reported by each class to avoid confusion about the overall model performance. In addition, standard metrics: precision, recall, and F1-measure were used to eliminate doubt about the model performance because accuracy is misleading for imbalance dataset. AUC values are also outlined due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets\cite{hu2015kernelized}, which are demonstrated by the ROC curves in Figure~\ref{fig:roc-curves}. NB shows the highest AUC values, achieved 0.978 for both cases while provides lowest classification results. On the other hand, KNN and SVM exhibit 0.972 and 0.835 AUCs when only lexical features are used; and 0.959 and 0.844 AUCs when a combination of lexical, punctuation mark and topic model-based features are used. Finally, LSTM demonstrates lowest AUC values among all machine learning models, achieved 0.82 AUC while GRU achieves 4.15\% higher AUC than LSTM. The conclusion drawn from the ROC curves also confirmed the robustness and superiority of KNN model for the segmentation of clinical exchange.

We observed the second highest performance of RNN, in particular, LSTM and GRU for the text segmentation. We believe that RNN will perform better if a large set of data is utilized. In this study, we employed 3,138 examples of boundary case which limit to achieve the best performance. We also observed that GRU performs better than LSTM which was observed in other previous study\cite{chung2014empirical}.

@@

Our proposed approach is novel for the segmentation of e-coaching text because previous studies mainly focus on the segmentation of text into sections, headers, and sentences in other medical domain. However, this study segmented clinical exchange into groups of MI behaviors which will significantly reduce the amount of resource and time required to segment clinical exchange manually. Furthermore, these segmentation models could be integrated with auto coding classifiers to build a software package of automatic coding procedure of clinical exchange.

The limitation of this study is that e-coaching text is collected from a single medical institute; formatting, style, and email segment can be different in other settings. Therefore, there is a need to replicate the experiments with different data sets. As our future work, we plan to evaluate our approach on other datasets involves in discourse analysis. We also plan to use more relevant features to improve model performance. For example, part-of-speech tagging\cite{hasan2016feedback} and distance from the beginning and end of the current sentence might significantly enhance the classification performance.

\section*{Conclusion}
Segmentation of e-coaching text is an integral part of developing an automated e-coaching intervention. Although several studies have done in clinical interventions, they are limited by the qualitative coding of clinical interactions. In addition, previous studies in the medical domain mainly segmented clinical text into sections and sentences, none of them are used for the segmentation of text into groups of MI behaviors in the setting of discourse analysis with email under the principle of motivational interviews. In this paper, we compared the performance of machine learning models for the task of segmentation of e-coaching text. We found out that k-nearest neighbor provides the best performance for the segmentation of text in terms of all performance metrics. Manual segmentation of e-coaching data is very resource-intensive and time-consuming task, which can significantly decrease the time and effort required to develop an effective behavioral intervention. Our proposed methods can help to identify individual text segments, which can be annotated directly with a classification model. This approach will also help for developing fully automated e-coaching and accelerate the pace of identifying effective communication strategies.

>>>>

\section*{Results}

Experimental results are reported with respect to ``new segment'' and ``same segment'' classes as well as their weighted average in Tables~\ref{tab:result_boundary}, \ref{tab:result_not_boundary}, and \ref{tab:result_weighted_avg}, respectively.\\

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM, KNN, and RNN methods for identification of ``new segment'' class in e-Coaching emails. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_boundary}
  \begin{tabular}{|l|l|l|l|p{0.15\linewidth}|p{0.15\linewidth}|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{lexical features only}} & \multicolumn{3}{|c|}{\textbf{lexical + punctuation (+ topic features except RNN)}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-measure} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-measure}\\ \hline

@@

As follows from Table~\ref{tab:result_boundary}, KNN outperforms all other models in terms of precision and F1-measure achieving 0.808 precision with 0.728 F1-measure when lexical features are used; 0.820 precision and 0.779 F1-measure when a combination of lexical, punctuation and topic features are used. NB demonstrates the lowest performance among all models in terms of all performance metrics. GRU has the second best results showing 10.68\% higher recall and 3.78\%  higher F1-measure than LSTM. On the other hand, SVM has the highest recall of 0.679 when only lexical features are used. When lexical features are used in combination with punctuation mark and topic model-based features, recall increases by 0.6\%, 2.5\%, and 11.92\%; and F1-measure increases by 0\%, 3.39\%, and 7\% for NB, SVM, and KNN models, respectively, while precision increases by 4.31\% and 1.49\% for SVM and KNN methods and decreases by 0.7\% for NB. \\

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM, KNN, and RNN methods for identification of ``same segment'' class in e-Coaching emails. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_not_boundary}
  \begin{tabular}{|l|l|l|l|p{0.15\linewidth}|p{0.15\linewidth}|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{lexical features only}} & \multicolumn{3}{|c|}{\textbf{lexical + punctuation (+ topic features except RNN)}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-measure} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-measure}\\ \hline

@@

Table~\ref{tab:result_not_boundary} summarizes the performance of NB, SVM, KNN, and RNN models for detecting ``same segment'' class. We observed that performance is remarkably better for ``same segment'' class compared to new segment detection, which is expected, since 96.71\% instances belong to ``same segment'' class. KNN achieves 22.40\%, 50.07\%, and 36.26\% higher precision, recall, and F1-measure, respectively, for lexical features; and 20.85\%, 33.96\%, and 27.47\% higher precision, recall, and F1-measure, respectively, for combined features compared to new segment classification. In contrast to boundary detection, RNN demonstrates the highest performance among all models. LSTM obtains 0.993 precision with 0.994 recall and F1-measure while GRU has 0.994 for all performance metrics. The impact of punctuation and topic model-based features is consistent with ``new segment'' classification. Results indicate that F1-measure increases by 0\%, and 0.1\% for SVM and KNN models and decreases by 0.1\% for NB. \\

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM, KNN, and RNN methods for segmentation of e-Coaching emails as a weighted average over ``same segment'' and ``new segment'' classification results. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_weighted_avg}
  \begin{tabular}{|l|l|l|l|p{0.15\linewidth}|p{0.15\linewidth}|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{lexical features only}} & \multicolumn{3}{|c|}{\textbf{lexical + punctuation (+ topic features except RNN)}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-measure} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-measure}\\ \hline

@@

Table~\ref{tab:result_weighted_avg} summarizes the weighted average results of the models for segmentation of e-Coaching emails. Overall, KNN has the best performance across all metrics and NB has the lowest performance among all methods. GRU demonstrates the same result as KNN for precision and F1-measure, but not for recall. SVM shows moderate performance, achieving 0.981, 0.982, and 0.983 for precision, recall, and F1-measure, respectively, when lexical features are used. Influence of the additional features is also consistent with the results in Tables~\ref{tab:result_boundary} and~\ref{tab:result_not_boundary}. Precision increases by 0.1\%, 0.2\%, and 0.3\%; recall increases by 0\%, 0.1\%, and 0.2\%; and F1-measure increases by 0\%, 0.2\%, and 0.3\% for NB, SVM, and KNN methods, respectively, when feature combination is used.\\

\section*{Discussion}
This study is the first effort to evaluate the automatic segmentation of e-Coaching emails. Experimental results indicate that KNN is the best model among all machine learning methods considered for this study. KNN achieved 0.986 F1-measure in overall, 0.779 and 0.993 F1-measures for detecting ``boundary'' and ``not boundary'', respectively. The robust performance of KNN provides the evidence that machine learning models are capable to learn conceptual information from clinical exchange. It also indicates that topic features are more important than lexical features, even when deep learning methods are employed. Although the domain of this study was intentionally quite small, we believe that our study is not limited to the e-coaching domain, and it can be successfully applied to other domains, in which discourse segmentation is an preliminary step for annotation.

Punctuation mark and topic model-based features made a significant improvement in performance of all machine learning methods. In nearly all cases, ML methods perform better, when lexical features are used in combination with punctuation and topic model-based features. This results also indicate that segmentation performance might be improved by adding additional relevant features.

@@

In this paper, results are reported by each class to avoid confusion regarding overall model performance due to severe class imbalance. In addition, standard metrics: precision, recall, and F1-measure were used to eliminate doubt about the model performance, since accuracy is misleading for imbalanced datasets. AUC is also utilized due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets\cite{hu2015kernelized}, which are demonstrated by the ROC curves in Figure~\ref{fig:roc-curves}. NB shows the highest AUC values, achieving 0.978 for both cases while provides lowest classification results. On the other hand, KNN and SVM exhibit 0.972 and 0.835 AUCs when only lexical features are used; and 0.959 and 0.844 AUCs when a combination of lexical, punctuation mark and topic model-based features are used. Finally, LSTM demonstrates lowest AUC values among all machine learning models, achieving 0.82 AUC, while GRU achieves 4.15\% higher AUC, than LSTM. The conclusion drawn from the ROC curves also confirmed the robustness and superiority of KNN model for the segmentation of clinical exchange.

We observed the second highest performance of RNN, in particular, LSTM and GRU for the text segmentation. We believe that RNN will perform better if a larger dataset is utilized. In this study, we employed 3,138 examples of boundary case which limit to achieve the best performance. We also observed that GRU performs better than LSTM, which was observed in previous studies\cite{chung2014empirical}.

@@

Our proposed approach is novel for the segmentation of e-Coaching emails, since previous studies mainly focused on the segmentation of text into sections, headers, and sentences. However, this study segmented clinical exchange into groups of MI behaviors which will significantly reduce the amount of resource and time required to segment clinical exchange manually. Furthermore, these segmentation models could be integrated with auto coding classifiers to build a software pipeline for automated annotation of exchanges.

The limitation of this study is that e-Coaching data is collected from a single medical institute; formatting, style, and email segment can be different in other settings. Therefore, there is a need to replicate the experiments with different data sets. As our future work, we plan to evaluate our approach on other datasets for discourse analysis.

%We also plan to use more relevant features to improve model performance. For example, part-of-speech tagging\cite{hasan2016feedback} and distance from %the beginning and end of the current sentence might significantly enhance the classification performance.

\section*{Conclusion}
Segmentation of e-coaching emails is an integral part of developing e-Coaching interventions. Although several studies have focused on clinical interventions, they are limited by the qualitative coding of clinical interactions. In addition, previous studies in the medical domain mainly segmented clinical text into sections and sentences, none of them considered segmentation of text into groups of MI behaviors in the setting of discourse analysis with emails. In this paper, we compared the performance of machine learning models for the task of segmentation of e-coaching text. We found out that k-nearest neighbor provides the best performance for the segmentation of text in terms of all performance metrics. Manual segmentation of e-coaching data is very resource-intensive and time-consuming task, which can significantly decrease the time and effort required to develop an effective behavioral intervention. Our proposed methods can help to identify individual text segments, which can be annotated directly with a classification model. This approach will also help for developing fully automated e-coaching and accelerate the pace of identifying effective communication strategies.
